---
title: "STA4813 Assignment 01"
author: "22692037_Tshepiso Mashiane"
date: "`r Sys.Date()`"
output:
  html_document: default
  #pdf_document: default
---

#Loading the Data
```{r}
library(readr)
Machinery_T <- read.csv("C:/Users/36076724/Downloads/Machinery_T.csv")
summary(Machinery_T)
head(Machinery_T)
tail(Machinery_T)
```

#Exemine the Data
```{r}
str(Machinery_T)
summary(Machinery_T)
```
#a. Constructing appropriate graphs and calculating appropriate summary statistics of both
#Average_Purchase_Price and Tractor_Power individually and then together. 
# plot the Histogram,Box plot and scatter plot
```{r}
library(gridExtra)
library(ggplot2)
library(dplyr)
# Average purchase price Histogram
average_purchace_price_hist<- ggplot(data=Machinery_T, aes(x=Average_Purchase_Price)) +
geom_histogram(binwidth = 100000, fill= "skyblue",color= "black") +
xlab("Average_Purchase_Price") + ylab("frequency") + labs(title = "(Figure 1(a) Histogram of Average_Purchase_Price )")
print(average_purchace_price_hist)

# Box plot of Average_Purchase_Price
average_purchase_box <- ggplot(data=Machinery_T, aes(y = Average_Purchase_Price)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(y = "Average Purchase Price (R)") +
  ggtitle("Figure 1(b) Box Plot of Average Purchase Price")
print(average_purchase_box)

# Tractor_power Histogram
Tractor_power_hist <- ggplot(data = Machinery_T, aes(x = Tractor_Power)) +
  geom_histogram(binwidth = 5,  fill= "skyblue",color= "black") +
  xlab("Tractor Power") + ylab("Frequency") + labs(title = "(Figure 2(a) Histogram Tractor Power)")
print(Tractor_power_hist)


# Box plot of Tractor_Power
tractor_power_box <- ggplot(Machinery_T, aes(y = Tractor_Power)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(y = "Tractor Power (kW)") +
  ggtitle("Figure 2(b) Box Plot of Tractor Power")
print(tractor_power_box)

#Calculate the Correlation between Tractor_Power and Average_Purchase_Price
correlation <- cor(Machinery_T$Average_Purchase_Price, Machinery_T$Tractor_Power)
print(correlation)

# Print the summary statistics
cat("Summary Statistics for Average Purchase Price:\n")
print(average_purchase_stats)
cat("\n")

cat("Summary Statistics for Tractor Power:\n")
print(tractor_power_stats)
cat("\n")
library(ggplot2)
# Create a combined scatter plot
combined_plot <- ggplot(Machinery_T, aes(x = Tractor_Power, y = Average_Purchase_Price)) +
  geom_point(color = "black") +
  labs(x = "Tractor Power (kW)", y = "Average Purchase Price (R)") +
  ggtitle("Combined Scatter Plot")
print(combined_plot)
```
#b.Fitting a simple linear regression model with Average_Purchase_Price as the response variable
#and Tractor_Power as the explanatory variable (Model 1).Also Interpreting the estimate slope coefficient
#β1.
# Fit a simple linear regression model
```{r}
Model_1 <- lm(Average_Purchase_Price ~ Tractor_Power, data = Machinery_T)
#summary(model_1)
summary(Model_1)
##need to interprete the slope

coef(summary(Model_1))
cat(" R squared= ",summary(Model_1)$r.squared, "\n",
    "Residual Standart Error= ",summary(Model_1)$sigma)

#Since the Intercept does not make sense, let subtract the mean from both the response variable and the independent variable
Machinery_T$Average_Purchase_Price<- Machinery_T$Average_Purchase_Price - mean(Machinery_T$Average_Purchase_Price)
Machinery_T$Tractor_Power<- Machinery_T$Tractor_Power- mean(Machinery_T$Tractor_Power)
Model_1a<- lm(Average_Purchase_Price~Tractor_Power, data= Machinery_T)
summary(Model_1a)
coef(summary(Model_1a))

#Are all assumptions of the linear regression of the met by the data?
# Obtain the fitted values and residuals
fitted_values <- fitted(Model_1)
residuals <- residuals(Model_1)
# Create the residual vs fitted values plot to assess the assumption of the constant varience of the residual
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual vs Fitted Values Plot")

##Linearity assessment:
#since the residuals are randomly scattered around the horizontal line at zero, it suggests that the linear regression mode_1 captures the linear relationship between the predictors and the response variable reasonably well. This indicates that the assumption of linearity is met.

##Homoscedasticity assessment:
#In the residual vs fitted values plot, we see a random scatter of points with approximately equal spread above and below zero. This indicates that the variance of the residuals is consistent across the range of fitted values/ predictor variables.

# Obtain the standardized residuals for normality assumption assessment
std_resid <- rstandard(Model_1)
# Create the Normal Q-Q plot
qqnorm(std_resid, xlab = "Theoretical Quantiles", ylab = "Standardized Residuals", main = "Normal Q-Q Plot")
qqline(std_resid)  

##since there is deviation from the line this indicate departures from normality, such as skewness or heavy tails.

```


# c.Creating a new variable logAverage_Purchase_Price, the natural log of Average_Purchase_Price and
#Fitting Model 2 where logAverage_Purchase_Price is now the response variable and Tractor_Power
#is still the explanatory variable. Also showing the estimated regression line equation.
```{r}
#since the Normality assumption is violated, we now need to transform the data.
# Create the new variable logAverage_Purchase_Price
Machinery_T$logAverage_Purchase_Price <- log(Machinery_T$Average_Purchase_Price)
Model_2 <- lm(logAverage_Purchase_Price ~ Tractor_Power, data = Machinery_T)
summary(Model_2)

coef(summary(Model_2))
cat(" R squared= ",summary(Model_2)$r.squared, "\n",
    "Residual Standart Error= ",summary(Model_2)$sigma)
#Since the Intercept does not make sense, let subtract the mean from both the response variable and the independent variable
#Machinery_T$logAverage_Purchase_Price<- Machinery_T$logAverage_Purchase_Price - mean(Machinery_T$logAverage_Purchase_Price)
#Machinery_T$Tractor_Power<- Machinery_T$Tractor_Power- mean(Machinery_T$Tractor_Power)
#Model_2a<- lm(Average_Purchase_Price~Tractor_Power, data= Machinery_T)
#summary(Model_2a)
#coef(summary(Model_2a))

####I am failing to pribt the fitted model!!~

```
#d.checking How does logAverage_Purchase_Price change when Tractor_Power increases by 1?
```{r}
#relating to the fitted model on c, The logAverage_purchse_price chances by R0.07888 when tractor power increase by 1

```

#e.
```{r}
#As Tractor_Power increases by 1, the logAverage_Purchase_Price changes by
#  log(Y)i= 10.96500600 + 0.07888368(1)=8 11.04388968
# therefore, Avarage_purchase_price by R57815.14242
#which is the exp(10.96500600)
```
#f.Are linear least squares regression (LLSR) assumptions satisfied in Model 2? Why or why
#not?
```{r}
# Obtain the fitted values and residuals
fitted_values <- fitted(Model_2)
residuals <- residuals(Model_2)
# Create the residual vs fitted values plot to assess the assumption of the constant varience of the residual
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual vs Fitted Values Plot")
##Linearity assessment:
#since the residuals are randomly scattered around the horizontal line at zero, it suggests that the linear regression mode_1 captures the linear relationship between the predictors and the response variable reasonably well. This indicates that the assumption of linearity is met.

##Homoscedasticity assessment:
#In the residual vs fitted values plot, we see a random scatter of points with approximately equal spread above and below zero. This indicates that the variance of the residuals is consistent across the range of fitted values/ predictor variables.

# Obtain the standardized residuals for normality assumption assessment
std_resid <- rstandard(Model_2)
# Create the Normal Q-Q plot
qqnorm(std_resid, xlab = "Theoretical Quantiles", ylab = "Standardized Residuals", main = "Normal Q-Q Plot")
qqline(std_resid)  

```

